---
title: "Searching and downloading species occurrences using `rgbif` R package"
author: "Javier Fernández-López"
date: "15/02/2022"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 1
    #code_folding: hide
  word_document: default
  pdf_document: default
---

<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE,warning=FALSE)
```
<br/><br/>
R is a language and environment for statistical computing and graphics, that allows for automation of repetitive task as well as improves the reproducibility of science. R is available as Free Software under the terms of the Free Software Foundation’s GNU General Public License in source code form.
<br/><br/>
This is a simple tutorial to learn how to use the `rgbif` R package to download, clean and visualize biodiversity occurrences from the [GBIF data portal](https://www.gbif.org/).

# Starting session

First, we install and load the R packages we will use. 

```{r rgbif1}
# Install packages (if needed)
#install.packages("rgbif")
#install.packages("ggplot2")
#install.packages("ggplot2")

# Load packages in R environment
library(rgbif)
library(ggplot2)
library(maps)
```
```{r rgbif0, echo=FALSE}
user <- "irec"
pwd <- "irecTest"
email <- "irec.aplicaciones@gmail.com"

occDlNeo <- occ_download(pred("taxonKey", 5218823), 
                         pred("hasCoordinate", TRUE),
                         pred("continent", "europe"), 
                         user = user, email = email, pwd = pwd)
```

<br/><br/>


# Selecting taxa name
There are some functions in the `rgbif` R package that are useful to look for scientific species names or other taxa that are included in the [GBIF Backbone Taxonomy](https://www.gbif.org/dataset/d7dddbf4-2cf0-4f39-9b2a-bb099caae36c). This is important when we don't know exactly the scientific name of a species, or when we don't know if the taxonomy of the species has changed in the last years. It is also important to check what is the *accepted* name in the GBIF backbone taxonomy. To do that we can use the `name_backbone` function.
<br/><br/>
Let's take the American mink, _Neovison vison_ as an example:

```{r rgbif2, message=FALSE, fig.height = 8, fig.width = 9.15, fig.align = "center"}
nBb <- name_backbone(name="Neovison vison", rank="species")
nBb
```
This table say that, even when _Neovison vison_ is the currently accepted species name by the scientific community, it is not the accepted name in the GBIF Backbone Taxonomy (see `SYNON...` in the `status` field). Thus, we should look for the list of synonyms of the accepted name in the GBIF Backbone Taxonomy. To do that, we take the `acceptedUsageKey` number and use the function `name_usage`.

```{r rgbif3, message=FALSE, fig.height = 8, fig.width = 9.15, fig.align = "center"}
# Accepted name in the GBIF Backbone Taxonomy
nBb$species
nBb$acceptedUsageKey
# Here we use the acceptedUsageKey to look for their synonyms
syn <- name_usage(key = 5218823, data = "synonyms")
syn
syn$data$scientificName
```
We see that there are three synonyms of _Mustela vison_ in the GBIF Backbone Taxonomy database: _Lutreola vison_, _Neovison vison_ and _Putorius vison_. We can check now the number of occurrences assigned at each name.

```{r rgbif4, message=FALSE, fig.height = 8, fig.width = 9.15, fig.align = "center"}
# Mustela vison
occ_count(5218823)
# Neovison vison
occ_count(2433652)
# Lutreola vison
occ_count(9309757)
# Putorius vison
occ_count(9458324)
```

In any case, we should use the accepted name in the GBIF Backbone Taxonomy database, since it contains all the occurrences from all synonyms. As we see, the name with the most number of occurrences is _Mustela vison_, so we will use its key number (5218823) for occurrence searches.
<br/><br/>

# Simple queries
As we already know, GBIF uses Darwin Core as the standard for biodiversity data storage. It means that if we don't specify the fields in which we are interested in, we will obtain a table with 84 fields (headers). Click [here](https://dwc.tdwg.org/terms/) for the list with all available fields in Darwin Core databases, and [here](https://www.gbif.org/es/data-quality-requirements-occurrences) for the required or strongly recommended fields in GBIF.


```{r rgbif5, message=FALSE, fig.height = 8, fig.width = 9.15, fig.align = "center"}
# Search with all Darwin Core fields
# We use limit to get only 10 records
occNeo <- occ_search(taxonKey = 5218823, limit = 10)
# Check the dimensions
dim(occNeo$data)

# Here we select only the fields we are interested in
selFields <- c("key","scientificName", "decimalLatitude","decimalLongitude",
               "issues", "country", "basisOfRecord","year")
# We use limit to get only 10 records
occNeo <- occ_search(taxonKey = 5218823, fields = selFields, limit = 10)
# Check the dimensions
dim(occNeo$data)
```

We will use just those fields for the rest of this tutorial. It is important to mention that `occ_search` function does not provide a DOI of our search. We'll see later how to obtain a DOI for our searches using `rgbif` package.
<br/><br/>

# Advanced queries and filtering

We can add different filter to our searches to refine our searches. For a complete list of available filters you can use `help(occ_search)`. Now we'll apply some of them that can be very useful to reduce the number of registers.<br/><br/>
Actually, we have already applied a first filter, called `limit` to reduce the maximum number of occurrence. The function `occ_search` allows a maximum of 100,000 occurrences, so if our taxa has more records we'll use a different approach (we'll come back about this later).<br/><br/>
Other useful filter can be the country. We can introduce the [country code](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-2) to filter for one or more countries.


```{r rgbif6, message=FALSE, fig.height = 8, fig.width = 9.15, fig.align = "center"}
# Look for occurrences in Spain and Italy
occNeo <- occ_search(taxonKey = 5218823, fields = selFields, country = c("ES","IT"))
summary(occNeo)
```

We can also delimit a geographical window for occurrence filtering. 

```{r rgbif7, message=FALSE, fig.height = 8, fig.width = 9.15, fig.align = "center"}
# Look for occurrences in a specific geographic window
occNeo <- occ_search(taxonKey = 5218823, fields = selFields, 
                     decimalLongitude = "1, 8", decimalLatitude = "48, 57")

# We can now map the region with the obtained occurrences
map("world", xlim = range(occNeo$data$decimalLongitude), ylim = range(occNeo$data$decimalLatitude))  
points(occNeo$data[ , c("decimalLongitude", "decimalLatitude")], pch = 19, col = "darkred", cex=0.5)

```

Many other filter such as the continent or collection year can be applied. For a complete list, please check `help(occ_search)`. We can then store the data in our computer in a CSV file.

```{r rgbif8, message=FALSE, fig.height = 8, fig.width = 9.15, fig.align = "center", eval=FALSE}
write.csv(occNeo$data, "Neovison_vison.csv")
```

<br/><br/>

# Submitting heavy queries

As we said, the function `occ_search` only allows to download 100,000 records from the GBIF data base. If our species of interest has more records, we should use `occ_download`. This function works in a similar way to queries in the GBIF web portal. It is necessary to [login](https://www.gbif.org/user/profile) with your account: user name, password and e-mail.

```{r rgbif9, message=FALSE, fig.height = 8, fig.width = 9.15, fig.align = "center", echo=FALSE}
user <- "irec"
pwd <- "irecTest"
email <- "irec.aplicaciones@gmail.com"
```

```{r rgbif10, message=FALSE, fig.height = 8, fig.width = 9.15, fig.align = "center", eval=FALSE}
user <- "irec"
pwd <- "******"
email <- "irec.aplicaciones@gmail.com"
```

The function `occ_download` submit a query to the GBIF portal. Once it has been processed, data will be ready for downloading and cleaning. Filters in `occ_download` can be applied by using the function `pred`. Check `help(pred)` for more information about the usage of filters in `occ_download`.

```{r rgbif11, message=FALSE, fig.height = 8, fig.width = 9.15, fig.align = "center", eval=FALSE}
# Submit a query
occDlNeo <- occ_download(pred("taxonKey", 5218823), 
                         pred("hasCoordinate", TRUE),
                         pred("continent", "europe"), 
                         user = user, email = email, pwd = pwd)
```

```{r rgbif12, message=FALSE, fig.height = 8, fig.width = 9.15, fig.align = "center"}
# Check the query data
occDlNeo

# Check the query status
occ_download_meta(occDlNeo)
```

It is important to highlight that `occ_download` provides us a DOI assigned to our query. As we already know, It will be useful for data citation and tracking.



